upwell_IDX

### 2020-04-30 upwell_IDX.Rmd
update

```{r}
# Installing and loading libraries
## devtools::install_github("robwschlegel/coastR")
## devtools::install_github("retostauffer/Rfoehnix")
library(tidyverse)
library(circular)
library(foehnix)
library(gridExtra)
library(geosphere)
library(heatwaveR)
library(coastR)
library(openair)
```

@@

# Selecting the important columns for each of the datasets
## Renaming the sites within the wind datasets to match the name of the sites at which seawater temperature was collected
## The wind data was obtained from the SAWS and the wind stations used were the closes stations to which temperature was collected
wind_fix <- function(df){
wind <- df %>%
  mutate(date = as.Date(as.character(date)),
         hour = as.numeric(as.character(hour)),
         dir = as.numeric(as.character(dir)),
         speed = as.numeric(as.character(speed)),
         temp_sites = case_when(station_name == "CAPE TOWN TABLE BAY" ~ "Sea Point",
                                station_name == "CAPE TOWN - ROYAL YACHT CLUB" ~ "Sea Point",
                                station_name  == "PORT NOLLOTH" ~"Port Nolloth",
                                station_name  == "CAPE TOWN SLANGKOP" ~ "Hout Bay",
                                station_name  == "LAMBERTSBAAI NORTIER" ~ "Lamberts Bay",
                                station_name  == "LANGEBAANWEG AWS" ~ "Saldanha Bay")) %>% 
  dplyr::select(temp_sites, date, hour, dir, speed) %>% 
  drop_na()
}

@@

wind_daily_tested <- wind_data %>% 
  filter(temp_sites %in% selected_sites) #%>%  # Matching the sites to wind temp sites
  # dplyr::group_by(temp_sites, date) 
# save(wind_daily_tested, file = "Data_coast_angle/wind_daily_tested.RData")
# The data provided by SAWS already reads wind from the direction which it is blowing
# RWS: Great find on this timeAverage function!
wind_daily <- timeAverage(wind_daily_tested, avg.time = "day", type = "temp_sites", vector.ws = TRUE)
save(wind_daily, file = "Data_coast_angle/wind_daily.RData")
```

@@

```{r}
# load("Data/wind_daily.RData") 
load("Data_coast_angle/wind_daily.RData") 
wind_daily <- wind_daily %>%
  dplyr::rename(sites = temp_sites,
                mean_speed = speed) %>% 
  drop_na()
load("Data/site_list_sub.Rdata")
load("Data/SACTN_US.RData")
```

@@

```{r}
# RWS: the cos() function works on radians, not degrees.
UI_angle <- wind_daily %>% # Making reference to the wind daily data created above
  dplyr::rename(site = sites) %>%  # Renaming sites to site
  #left_join(south_africa_away_wide[,c(2,21)], by = "site") %>%  # Joining wind data to South Africa wide data column 2 and 21. by site 
  drop_na() %>% # removing the na values
  #dplyr::rename(coast_angle = heading) %>% # Renaming the column heading to coast_angle
  mutate(ui.saws = mean_speed * (cos(deg_rad(dir - 160)))) %>% # This is the orientation of the coastal angle for the west coast
  dplyr::rename(t = date) %>% # renaming date to t
  drop_na()
# save(UI_angle, file = "Data_coast_angle/UI_angle.RData")
# save(UI_angle, file = "Data/UI_angle.RData")
```

@@

```{r}
# Loading the insitu temperature data along the wc
load("Data/SACTN_US.RData") # temperature data for all the sites with a 30year time series
# exceedance <-
#   function(data,
#            x = t,
#            y = temp,
#            threshold,
#            below = FALSE,
#            minDuration = 5,
#            joinAcrossGaps = TRUE,
#            maxGap = 2,
#            maxPadLength = 3)
# Upwelling temperature threshold 12.4 / 30th percentile
SACTN_upwell <- UI_angle %>% 
  dplyr::rename(temp = ui.saws) %>%
  group_by(site) %>%
  mutate(t = as.Date(t)) %>% 
  # mutate(min_t = min(t), 
  #        max_t = max(t)) %>% 
  nest() %>% # apply the following functions to all of the variables in the dataset
  # RWS: Please stop using uneven periods for your climatology periods 12 month periods should be used.
  mutate(clim = purrr::map(data, ts2clm, climatologyPeriod = c("1992-07-09", "2017-12-21")), # creating a column of climatologies. Column will be named clim
         # NB: A threshold of 3 appeared to be far to strict
         # purr::map - applies a function to each element of a vector
         exceed = purrr::map(clim, exceedance, minDuration = 1, threshold = 1)) %>%  # Upwelling cannot be described as an event. Upwelling can last for a few hours. Given that we have daily data, upwelling events minimum duration here will be 1day
  # Detect consecutive days in exceedance of a given threshold.
  # mutate() %>% 
  select(-data, -clim) %>% 
  unnest(cols = exceed) %>%
  filter(row_number() %% 2 == 1) %>%
  unnest(cols = exceed) %>% # creates a column for each variables
  dplyr::rename(ui.saws = temp) %>% # rename upwelling index vale to temp so that it could work with the function
  select(site, t, ui.saws, exceedance) # selecting only these variables

# RWS: If with the above code your primary objective is to determine when ui.saws is above 1, that may be done with the following code
SACTN_upwell <- UI_angle %>% 
  mutate(exceedance = ifelse(ui.saws >= 1, TRUE, FALSE),
         t = as.Date(t))

# Calculate quantiles of upwelling index

@@

# SACTN_upwell_transform <- as.data.frame(read.zoo(transform( SACTN_upwell, Date = as.POSIXct(t) ), FUN = identity ))
# There are a few ideas on how to go about doing this
# The first is to calculate rolling climatology and to check the upwelling phenology against that
# in order to determine changes in upwelling values, but this won't give us the metrics we want
# Rather what we can do is set a static bottom threshold below which we are interested in looking for upwelling
# We then use the upwelling index values created above as a second filter to determine the metrics of the upwelling
# Unfortunately the way that the heatwaveR functions work well does with purrr. 
# I've figured this out once before but can't remember where now.
# Rather than go through the process I have rather just made a convenience function below that
# does what I want and can simply be given to purrr::map() without any faffing about.

# Detect event: 
detect_event_custom <- function(df){
  res <- detect_event(df, threshClim2 = df$exceedance, minDuration = 1, coldSpells = T)
  return(res)
}
# Calculate the upwelling event metrics
SACTN_upwell_base <- SACTN_US %>% 
  dplyr::rename(t = date) %>% 
  group_by(site) %>% 
  nest() %>% 
  # The climatology base period used here is up for debate...
  # The choice of the 30th percentile threshold also needs to be justified and sensitivity tested
  mutate(clim = purrr::map(data, ts2clm, pctile = 25, climatologyPeriod = c("1992-07-09", "2005-10-19"))) %>%
  select(-data) %>% 
  unnest(cols = clim) %>%
  left_join(SACTN_upwell, by = c("site", "t")) %>%
  filter(!is.na(exceedance)) %>%
  group_by(site) %>% 
  # mutate(thresh = mean(seas, na.rm = T)-sd(seas, na.rm = T)) %>% # Manually set threshold to the mean
  # mutate(thresh = 10) %>% # Manually set threshold to a static value. Doesn't work across all sites
  # mutate(thresh = quantile(temp, 0.3, na.rm = T)) %>% # Manually set threshold to a single quantile
  nest() %>% 
  mutate(exceed = purrr::map(data, detect_event_custom)) %>% 
  select(-data) #%>% 
  # unnest() %>% 
  # filter(row_number() %% 2 == 0) %>% # Select event summary metrics
  # filter(row_number() %% 2 == 1) %>% # Select daily values
  # unnest()
# Unpack the event metric reults
SACTN_upwell_events <- SACTN_upwell_base %>% 
  unnest() %>%
  filter(row_number() %% 2 == 0) %>%
  unnest()
# save(SACTN_upwell_events, file = "Data/SACTN_upwell_events.RData")
# save(SACTN_upwell_events, file = "Data_coast_angle/SACTN_upwell_events.RData")
# Unpack the daily climatology results
SACTN_upwell_clims <- SACTN_upwell_base %>% 
  unnest() %>%
  filter(row_number() %% 2 == 1) %>% 
  unnest()
# save(SACTN_upwell_clims, file = "Data/SACTN_upwell_clims.RData")
# The above chunk of code appears to work as expected.
# What needs to be done now is that the thresholds for duration and UI strength need to be justified/decided uopn. 
    # Min duration for an upwelling event - 1day and threshold -1
# What also needs to be decided is if we are interested in any upwelling results throughout the year,
# or only during the upwelling season, whenever that may be for each site.
# I think we have demonstrated the technical capability needed to answer the question of whether or not upwelling
# is changing over time.
# Now we really need to focus on which of these parameters need to best be tweaked to answer that question.
```

@@

```{r}
# match_func <- function(df){
#   match <- SACTN_MCS_clims  %>% 
#   left_join(df, by = c("site", "t")) %>% 
#   na.trim()
#   return(match)
# }
# 
# index <- index %>% 
#   dplyr::rename(site = sites) %>% 
#   dplyr::rename(t = date)
# 
# match <- match_func(df = index)
# 
# upwell_MCS <- ggplot(data = match, aes(x = t, y = ui.saws)) +
#   geom_line() +
#   geom_point(data = filter(match, event == TRUE), colour = "navy") +
#   geom_rug(data = filter(match, saws.condition == "upwelling"), sides = "b") +
#   scale_x_date(expand = c(0, 0)) +
#   facet_wrap(~site, ncol = 1) +
#   theme_grey() +
#   labs(y = "Temperature (°C)", x = NULL)
# upwell_MCS
# Plot some of the Upwelling/MCS results
SP_upwelling <- SACTN_upwell_clims %>% 
  filter(site == "Sea Point", lubridate::year(t) %in% 2016:2018)
plot1 <- ggplot(SP_upwelling, aes(x = t, y = temp)) +
  geom_flame(aes(y = thresh, y2 = temp), fill = "navy", n = 3) +
  geom_point(data = filter(SP_upwelling, !is.na(event_no))) +
  geom_line(aes(y = thresh)) +
  geom_line(alpha = 0.7) +
  geom_rug(data = filter(SP_upwelling, event_no > 0), sides = "b")
plot1

# The dots on the plot help to show the length of each wind driven upwelling event detected.
# This supports the use of this methodology.
# Up next is fitting linear models to the upwelling metric results to see what trends shake out.
# Linear Models
lm_coeff <- function(df){
  res <- lm(formula = val ~ date_peak, data = df)
  res_coeff <- round(as.numeric(res$coefficient[2]), 4)
}
# Changes in upwelling metrics
SACTN_upwell_lm <- SACTN_upwell_events %>% 
  select(-c(index_start:index_end)) %>% 
  gather(key = "var", value = "val", -c(site:date_end)) %>% 
  group_by(site, var) %>% 
  nest() %>% 
  mutate(slope = purrr::map(data, lm_coeff)) %>% 
  select(-data) %>% 
  unnest() %>% 
  # convert from daily to decadal values
  mutate(slope = slope * 365.25*10)
```

>>>>

```{r}
# Installing and loading libraries
## devtools::install_github("robwschlegel/coastR")
## devtools::install_github("retostauffer/Rfoehnix")
library(tidyverse)
library(circular)
#library(foehnix)
library(gridExtra)
library(geosphere)
library(heatwaveR)
library(coastR)
library(openair)
```

@@

wind_fix <- function(df){
wind <- df %>%
  mutate(date = as.Date(as.character(date)),
         hour = as.numeric(as.character(hour)),
         dir = as.numeric(as.character(dir)),
         speed = as.numeric(as.character(speed)),
         temp_sites = case_when(station_name == "CAPE TOWN TABLE BAY" ~ "Sea Point",
                                station_name == "CAPE TOWN - ROYAL YACHT CLUB" ~ "Sea Point",
                                station_name  == "PORT NOLLOTH" ~"Port Nolloth",
                                station_name  == "CAPE TOWN SLANGKOP" ~ "Hout Bay",
                                station_name  == "LAMBERTSBAAI NORTIER" ~ "Lamberts Bay",
                                station_name  == "LANGEBAANWEG AWS" ~ "Saldanha Bay")) %>% 
  dplyr::select(temp_sites, date, hour, dir, speed) %>% 
  drop_na()
}

@@

wind_daily_tested <- wind_data %>% 
  filter(temp_sites %in% selected_sites) 
# save(wind_daily_tested, file = "Data_coast_angle/wind_daily_tested.RData")

wind_daily <- timeAverage(wind_daily_tested, avg.time = "day", type = "temp_sites", vector.ws = TRUE)
save(wind_daily, file = "Data_coast_angle/wind_daily.RData")
```

@@

```{r}
load("Data_coast_angle/wind_daily.RData") 
wind_daily <- wind_daily %>%
  dplyr::rename(sites = temp_sites,
                mean_speed = speed) %>% 
  drop_na()
load("Data/site_list_sub.Rdata")
load("Data/SACTN_US.RData")
```

@@

```{r}
UI_angle <- wind_daily %>% 
  dplyr::rename(site = sites) %>%  
  drop_na() %>% 
  mutate(ui.saws = mean_speed * (cos(deg_rad(dir - 160)))) %>% # This is the orientation of the coastal angle for the west coast
  dplyr::rename(t = date) %>% # renaming date to t
  drop_na()
# save(UI_angle, file = "Data_coast_angle/UI_angle.RData")
# save(UI_angle, file = "Data/UI_angle.RData")
```

@@

```{r}
# Loading the insitu temperature data along the wc
load("Data/SACTN_US.RData")
SACTN_upwell <- UI_angle %>% 
  dplyr::rename(temp = ui.saws) %>%
  group_by(site) %>%
  mutate(t = as.Date(t)) %>% 
  # mutate(min_t = min(t), 
  #        max_t = max(t)) %>% 
  nest() %>% 
  mutate(clim = purrr::map(data, ts2clm, climatologyPeriod = c("1993-01-01", "2017-12-21")), 
         exceed = purrr::map(clim, exceedance, minDuration = 1, threshold = 1)) %>%  
  select(-data, -clim) %>% 
  unnest(cols = exceed) %>%
  filter(row_number() %% 2 == 1) %>%
  unnest(cols = exceed) %>% 
  dplyr::rename(ui.saws = temp) %>%
  select(site, t, ui.saws, exceedance)

# OR
SACTN_upwell <- UI_angle %>% 
  mutate(exceedance = ifelse(ui.saws >= 1, TRUE, FALSE),
         t = as.Date(t))x

@@

# Detect event: 
detect_event_custom <- function(df){
  res <- detect_event(df, threshClim2 = df$exceedance, minDuration = 1, coldSpells = T)
  return(res)
}
# Calculate the upwelling event metrics
SACTN_upwell_base <- SACTN_US %>% 
  dplyr::rename(t = date) %>% 
  group_by(site) %>% 
  nest() %>% 
  mutate(clim = purrr::map(data, ts2clm, pctile = 25, climatologyPeriod = c("1993-01-01", "2005-10-19"))) %>%
  select(-data) %>% 
  unnest(cols = clim) %>%
  left_join(SACTN_upwell, by = c("site", "t")) %>%
  filter(!is.na(exceedance)) %>%
  group_by(site) %>% 
  nest() %>% 
  mutate(exceed = purrr::map(data, detect_event_custom)) %>% 
  select(-data)

SACTN_upwell_events <- SACTN_upwell_base %>% 
  unnest() %>%
  filter(row_number() %% 2 == 0) %>%
  unnest()
# save(SACTN_upwell_events, file = "Data/SACTN_upwell_events.RData")
# save(SACTN_upwell_events, file = "Data_coast_angle/SACTN_upwell_events.RData")

SACTN_upwell_clims <- SACTN_upwell_base %>% 
  unnest() %>%
  filter(row_number() %% 2 == 1) %>% 
  unnest()
# save(SACTN_upwell_clims, file = "Data/SACTN_upwell_clims.RData")
```

@@

```{r}
SP_upwelling <- SACTN_upwell_clims %>% 
  filter(site == "Sea Point", lubridate::year(t) %in% 2016:2018)
plot1 <- ggplot(SP_upwelling, aes(x = t, y = temp)) +
  geom_flame(aes(y = thresh, y2 = temp), fill = "navy", n = 3) +
  geom_point(data = filter(SP_upwelling, !is.na(event_no))) +
  geom_line(aes(y = thresh)) +
  geom_line(alpha = 0.7) +
  geom_rug(data = filter(SP_upwelling, event_no > 0), sides = "b")
plot1

# Linear Models
lm_coeff <- function(df){
  res <- lm(formula = val ~ date_peak, data = df)
  res_coeff <- round(as.numeric(res$coefficient[2]), 4)
}
# Changes in upwelling metrics
SACTN_upwell_lm <- SACTN_upwell_events %>% 
  select(-c(index_start:index_end)) %>% 
  gather(key = "var", value = "val", -c(site:date_end)) %>% 
  group_by(site, var) %>% 
  nest() %>% 
  mutate(slope = purrr::map(data, lm_coeff)) %>% 
  select(-data) %>% 
  unnest() %>% 
  # convert from daily to decadal values
  mutate(slope = slope * 365.25*10)
```
